---
title: Simulations - phylogenetics
---

# extract top 100 sequences

```{sh}
WD=Simulations/03_phylogenetics
cd $WD

#MSA
cp Simulations/01_create_sequences/sequences10bases.fasta .


grep -c ">"  sequences10bases.fasta #1048576
head -n200 sequences10bases.fasta > MSAtop100.fasta

#v2.2.0
seqkit stat MSAtop100.fasta

```
file             format  type  num_seqs  sum_len  min_len  avg_len  max_len
MSAtop100.fasta  FASTA   DNA        100    1,000       10       10       10        948  28,375,536   29,932   29,932   29,932

# 2. phylogenetic tree computation
```{sh}
conda activate iqtree_2.1.2  
iqtree -s  MSAtop100.fasta -pre iqtree   -m TIM2+F+I -nt 60
```
Optimal log-likelihood: -254.570
Rate parameters:  A-C: 0.00124  A-G: 0.00010  A-T: 0.00124  C-G: 1.00000  C-T: 0.39326  G-T: 1.00000
Base frequencies:  A: 0.749  C: 0.065  G: 0.069  T: 0.117
Proportion of invariable sites: 0.565
Parameters optimization took 9 rounds (0.038 sec)
BEST SCORE FOUND : -254.570
Total tree length: 133.667

Total number of iterations: 470
CPU time used for tree search: 20.286 sec (0h:0m:20s)
Wall-clock time used for tree search: 25.524 sec (0h:0m:25s)
Total CPU time used: 89.287 sec (0h:1m:29s)
Total wall-clock time used: 30.505 sec (0h:0m:30s)

Analysis results written to:
  IQ-TREE report:                iqtree.iqtree
  Maximum-likelihood tree:       iqtree.treefile
  Likelihood distances:          iqtree.mldist
  Screen log file:               iqtree.log

 ------------------------------------------

# moving to R 
# R comparison of asymdist 1D vs. phylogenetic distances 

```{r}
options(digits=10)
capasydisData <- read.csv("Simulations/02_build_axes/output_build_axesv0.1.8_R1_R2_R3/output_R1_R2_R3_with_color.csv",header=TRUE)

head(capasydisData)
require(dplyr)
```

### how does capasydis compare to a "classical" phylogenetic analysis?
```{r}
library(ape)

# Read the mldist file
df <- read.table("iqtree/iqtree.mldist", header = FALSE, skip = 1, stringsAsFactors = FALSE)

# Extract the sequence names
seq_names <- df[, 1]

# Create a distance matrix
dist_matrix <- as.dist(df[, -1])

# Assign sequence names to the matrix
attr(dist_matrix, "Labels") <- seq_names
hc <-hclust(dist_matrix) 

require(ape)
plot(as.phylo(hc), cex = 0.6, label.offset = 0.1)
png("M10bases_phylogram.png",width = 30,height = 20,units="cm",res=300)
plot(as.phylo(hc), type = "phylogram", show.tip.label = TRUE, 
     edge.color = "black", edge.width = 1, edge.lty = 1,  
     tip.color = "black")
dev.off()
# Unrooted
png("M10bases_unrooted.png",width = 10,height = 10,units="cm",res=300)
plot(as.phylo(hc), type = "unrooted", cex = 0.6,
     no.margin = TRUE)
dev.off()

# Convert to a regular matrix
matrix_form <- as.matrix(dist_matrix)
dim(matrix_form)
matrix_form[1:3,1:3]


png("M10bases_matrix.png",width = 10,height = 10,units="cm",res=300)
plot(matrix_form,pch=16,xlim=c(0,12),ylim=c(0,10))    
text(matrix_form,label=seq_names,pos=4,cex=0.4)
dev.off()

png("M10bases_matrix_zoom_bottom.png",width = 10,height = 10,units="cm",res=300)
plot(matrix_form,pch=16,xlim=c(8,10),ylim=c(0,3))    
text(matrix_form,label=seq_names,pos=4,cex=0.4)
dev.off()


```

# How do the distances compare?
if we take the first sequence as reference, we can compare the phylogenetic distances to the asymmetrical distances in 1D (asymDist1D) for each sequence.
```{r}
Phylo1D <- matrix_form[,1]

# capasydisData checking
table(capasydisData[1:100,]$label == names(Phylo1D)) #OK


ComparisonTable <- cbind(capasydisData[1:100,],phyloDist=Phylo1D)
head(ComparisonTable)

# getting the number of mutations for each sequence
library(tidyverse)
ComparisonTable <- ComparisonTable %>% 
  mutate(
    prefix = str_sub(label, 1, 2),
    Nmut  = str_remove(prefix, "^S")
  )
ComparisonTable <- ComparisonTable %>%
  mutate(
    prefix = str_sub(label, 1, 2),
    Nmut  = case_when(
      prefix == "S_" ~ 10L,
      TRUE           ~ as.numeric(str_remove(prefix, "^S"))
    )
)

require(ggplot2)

Cols_unique_pairs <- unique(ComparisonTable[c("color", "Nmut")])

png("Top100seq_asymdist_xVsPhyloDist.png",width = 10,height = 10,units="cm",res=300)
ComparisonTable %>% ggplot(aes(y=x,x=phyloDist,label=label,color=color))+ 
    geom_point()+ geom_text(size=1.5,hjust = -0.3, nudge_x = 0.03) + 
    xlab("phyloDist (mldist)") +
    ylab("CAPASYDIS x axis") +
    xlim(0,12)+
      theme_light()+
  scale_fill_identity(
    name = "Color",
    breaks = Cols_unique_pairs$color,
    labels = Cols_unique_pairs$Nmut,
    guide = "legend"
  ) +
  scale_color_identity(
    name = "Number of mutations",
    breaks = Cols_unique_pairs$color,
    labels = Cols_unique_pairs$Nmut,
    guide = "legend"
  ) 
dev.off()
#zooming in on the bottom right corner
png("Top100seq_asymdist_xVsPhyloDist_zoom.png",width = 10,height = 10,units="cm",res=300)
ComparisonTable %>% ggplot(aes(y=x,x=phyloDist,label=label,color=color))+ 
    geom_point()+ geom_text(size=1.5,hjust = -0.3, nudge_x = 0.03) + 
    xlab("phyloDist (mldist)") +
    ylab("CAPASYDIS x axis") +
    xlim(8.5,9.25)+
      theme_light()+
  scale_fill_identity(
    name = "Color",
    breaks = Cols_unique_pairs$color,
    labels = Cols_unique_pairs$Nmut,
    guide = "legend"
  ) +
  scale_color_identity(
    name = "Number of mutations",
    breaks = Cols_unique_pairs$color,
    labels = Cols_unique_pairs$Nmut,
    guide = "legend"
  ) 
dev.off()
```

**Numerical investigations**
```{r}
ComparisonTable $phyloDist %>% summary()
```
   Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
0.000000 8.999999 9.000000 8.892161 9.000000 9.000000 

```{r}
ComparisonTable $phyloDist %>% unique()
```  
[1] 0.0000000 8.7451681 8.9999987 8.9999982 8.9999989 9.0000000
so only 6 values (see plot "Top100seq_asymdist_xVsPhyloDist_zoom.png")

Counting the number of sequences for each phyloDist value:
```{r}
ComparisonTable %>%  group_by(phyloDist) %>% summarise(Nseq = n(),Percent = n()/nrow(ComparisonTable)*100)
```
  phyloDist  Nseq Percent
      <dbl> <int>   <dbl>
1      0        1       1
2      8.75     7       7
3      9.00    17      17
4      9.00    18      18
5      9.00     3       3
6      9       54      54

1      0.0000000     1       1
2      8.7451681     7       7
3      8.9999987    17      17
4      8.9999982    18      18
5      8.9999989     3       3
6      9.0000000    54      54
  

# plotting phyloDist vs. Nmut
```{r}
png("Top100seq_Nmut_VsPhyloDist.png",width = 10,height = 10,units="cm",res=300)
ComparisonTable %>% ggplot(aes(x=Nmut,y=phyloDist,color=color,label=label))+ 
    geom_point()+ geom_text(size=1.5,hjust = -0.3, nudge_x = 0.03) + 
    xlab("Number of mutations") +
    ylab("phyloDist (mldist)") +
    theme_light()+
    ylim(8.5,9.5)+
    xlim(0,5)+
    theme(legend.position = "none")

dev.off()
```
